---
title: "GPHS 782: USGS Streamflow and Water-Quality Data Lab"
subtitle: "Using the modern USGS Water Data APIs in R"
author: "GPHS 782 Watershed Science Seminar"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Learning objectives

In this lab, we will treat stream gages as windows into **integrated watershed behavior**.

You will learn how to:

1.  clone the course GitHub repository;
2.  locate monitoring stations in USGS Water Data for the Nation;
3.  query the modern USGS Water Data APIs from R;
4.  download discharge, water temperature, and turbidity data;
5.  calculate annual water yield;
6.  compare turbidity among watersheds; and
7.  evaluate whether differences in data availability affect the conclusions we can draw.

# Clone the GitHub repository

Our course repository is:

<https://github.com/kellyloria/GPHS_782_watershed_sci_seminar.git>

## Using RStudio

Go to:

**File → New Project → Version Control → Git**

Then enter:

``` text
https://github.com/kellyloria/GPHS_782_watershed_sci_seminar.git
```

Choose a location on your computer and click **Create Project**.

## Using the terminal

``` bash
git clone https://github.com/kellyloria/GPHS_782_watershed_sci_seminar.git
cd GPHS_782_watershed_sci_seminar
```

**Checkpoint:** Look in the Files pane in RStudio. Are you working inside the cloned repository?

# Explore USGS Water Data

Before writing code, open:

- USGS Water Data for the Nation: <https://waterdata.usgs.gov/>
- USGS Water Data APIs: <https://www.usgs.gov/tools/usgs-water-data-apis>
- USGS API landing page: <https://api.waterdata.usgs.gov/>
- `dataRetrieval` documentation: <https://water.code-pages.usgs.gov/dataRetrieval/>

The `dataRetrieval` package provides R functions that construct requests to the modern USGS Water Data APIs.

For this class, we will use functions beginning with:

``` text
read_waterdata_*
```

# Install and load packages

Run these installation lines once if needed.

```{r packages, eval=FALSE}
install.packages(c(
  "dataRetrieval",
  "tidyverse",
  "lubridate",
  "scales"
))
```

Load packages:

```{r libraries}
library(dataRetrieval)
library(tidyverse)
library(lubridate)
library(scales)
```

# Our watersheds

We will compare two Lake Tahoe tributaries with the much larger Truckee River.

```{r sites}
sites <- tribble(
  ~site_name,        ~monitoring_location_id, ~drainage_area_mi2,
  "Blackwood Creek", "USGS-10336660",          11.2,
  "Incline Creek",   "USGS-10336700",           6.74,
  "Truckee River",   "USGS-10348000",        1067
)

sites
```

The sites differ greatly in drainage area, so **total discharge volume** and **watershed-normalized runoff** answer different questions.

# Look up station metadata with the API

```{r site-metadata}
site_metadata <- read_waterdata_monitoring_location(
  monitoring_location_id = sites$monitoring_location_id
)

site_metadata %>%
  as_tibble() %>%
  select(
    any_of(c(
      "id",
      "monitoring_location_id",
      "monitoring_location_name",
      "state_name",
      "county_name",
      "hydrologic_unit_code",
      "drainage_area"
    ))
  )
```

# USGS parameter codes

```{r parameter-codes}
parameter_codes <- tribble(
  ~parameter_code, ~variable,
  "00060", "Discharge",
  "00010", "Water temperature",
  "63680", "Turbidity"
)

parameter_codes
```

Ask what the parameter codes mean:

```{r parameter-metadata}
parameter_metadata <- read_waterdata_parameter_codes(
  parameter_code = parameter_codes$parameter_code
)

parameter_metadata %>%
  as_tibble() %>%
  select(any_of(c(
    "parameter_code",
    "parameter_name",
    "parameter_description",
    "unit_of_measure"
  )))
```

# Check data availability before downloading

A common mistake is assuming that every USGS gage measures every variable.

```{r time-series-meta}
ts_metadata <- read_waterdata_ts_meta(
  monitoring_location_id = sites$monitoring_location_id
) %>%
  as_tibble()

names(ts_metadata)
```

Filter for the variables we care about:

```{r availability}
availability <- ts_metadata %>%
  filter(parameter_code %in% parameter_codes$parameter_code) %>%
  left_join(parameter_codes, by = "parameter_code") %>%
  left_join(sites, by = "monitoring_location_id")

availability %>%
  select(
    any_of(c(
      "site_name",
      "monitoring_location_id",
      "variable",
      "parameter_code",
      "computation_period_identifier",
      "begin",
      "end",
      "begin_date",
      "end_date",
      "unit_of_measure"
    ))
  ) %>%
  arrange(site_name, variable)
```

Because metadata fields can evolve, `any_of()` keeps the display step from failing if a column name changes.

# Define a water year

We will compare **Water Year 2024**.

```{r dates}
start_date <- as.Date("2023-10-01")
end_date   <- as.Date("2024-09-30")

start_date
end_date
```

Water Year 2024 begins October 1, 2023 and ends September 30, 2024.

# Download daily mean discharge

For discharge, we do not need every 15-minute observation.

- Parameter `00060` = discharge
- Statistic `00003` = mean

```{r download-discharge}
q_daily <- read_waterdata_daily(
  monitoring_location_id = sites$monitoring_location_id,
  parameter_code = "00060",
  statistic_id = "00003",
  time = c(start_date, end_date)
) %>%
  as_tibble()

glimpse(q_daily)
```

Clean the discharge data:

```{r clean-discharge}
q_daily_clean <- q_daily %>%
  transmute(
    monitoring_location_id,
    date = as.Date(time),
    discharge_cfs = as.numeric(value)
  ) %>%
  left_join(sites, by = "monitoring_location_id")

head(q_daily_clean)
```

# Check discharge completeness

```{r q-completeness}
q_completeness <- q_daily_clean %>%
  group_by(site_name) %>%
  summarise(
    first_date = min(date, na.rm = TRUE),
    last_date = max(date, na.rm = TRUE),
    days_with_data = n_distinct(date[!is.na(discharge_cfs)]),
    missing_q_values = sum(is.na(discharge_cfs)),
    .groups = "drop"
  )

q_completeness
```

### Question

Would you calculate annual water yield if a site were missing several months of discharge data? Why or why not?

# Plot the hydrographs

```{r hydrographs, fig.width=10, fig.height=6}
ggplot(q_daily_clean, aes(x = date, y = discharge_cfs)) +
  geom_line() +
  facet_wrap(~ site_name, scales = "free_y", ncol = 1) +
  labs(
    x = NULL,
    y = expression("Daily mean discharge (ft"^3*"/s)"),
    title = "Water Year 2024 hydrographs"
  ) +
  theme_minimal()
```

Consider controls on hydrograph shape:

- watershed size,
- snow accumulation and melt,
- groundwater contributions,
- channel storage,
- reservoirs and regulation,
- diversions,
- elevation,
- precipitation timing.

# Calculate total annual streamflow volume

Daily discharge is reported in cubic feet per second.

$$V_{day} = Q \times 86,400 \; \frac{seconds}{day}$$

and

$$1 \; acre\!-\!foot = 43,560 \; ft^3$$

Therefore:

```{r conversion}
cfs_day_to_acre_feet <- 86400 / 43560
cfs_day_to_acre_feet
```

Approximately 1 cfs sustained for one day equals 1.9835 acre-feet.

```{r annual-volume}
annual_yield <- q_daily_clean %>%
  mutate(
    daily_volume_af = discharge_cfs * cfs_day_to_acre_feet
  ) %>%
  group_by(site_name, drainage_area_mi2) %>%
  summarise(
    days_with_q = sum(!is.na(discharge_cfs)),
    annual_volume_af = sum(daily_volume_af, na.rm = TRUE),
    .groups = "drop"
  )

annual_yield
```

**Important:** `na.rm = TRUE` does not fill missing days. Interpret annual volume together with `days_with_q`.

# Normalize annual flow by watershed area

One square mile contains 640 acres.

```{r runoff-depth}
annual_yield <- annual_yield %>%
  mutate(
    watershed_area_acres = drainage_area_mi2 * 640
  )

annual_yield
```

Plot total volume:

```{r yield-plots, fig.width=9, fig.height=5}
ggplot(annual_yield, aes(x = reorder(site_name, annual_volume_af), y = annual_volume_af)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(
    x = NULL,
    y = "Annual discharge volume (acre-feet)",
    title = "Total annual streamflow volume"
  ) +
  theme_minimal()
```

### Questions

1.  Which site has the largest **total** annual discharge?
2.  Which site has the largest **area-normalized runoff depth**?
3.  Why are these rankings not necessarily the same?
4.  Is runoff depth a direct measurement of precipitation? Why not?

# Download high-frequency water temperature and turbidity

Use the continuous-values API for sensor observations.

```{r download-continuous}
sensor_data <- read_waterdata_continuous(
  monitoring_location_id = sites$monitoring_location_id,
  parameter_code = c("00010", "63680"),
  time = c(start_date, end_date)
) %>%
  as_tibble()

glimpse(sensor_data)
```

This request may take longer because continuous records contain many observations.

# Clean the sensor data

```{r clean-sensors}
sensor_clean <- sensor_data %>%
  transmute(
    monitoring_location_id,
    datetime = as.POSIXct(time),
    parameter_code,
    value = as.numeric(value),
    unit_of_measure = unit_of_measure
  ) %>%
  left_join(parameter_codes, by = "parameter_code") %>%
  left_join(sites, by = "monitoring_location_id")

sensor_clean %>%
  count(site_name, variable)
```

# Summarize water temperature

Calculate daily mean temperature:

```{r daily-temp}
temp_daily <- sensor_clean %>%
  filter(parameter_code == "00010") %>%
  mutate(date = as.Date(datetime)) %>%
  group_by(site_name, date) %>%
  summarise(
    mean_temp_c = mean(value, na.rm = TRUE),
    n_obs = sum(!is.na(value)),
    .groups = "drop"
  )

temp_daily
```

### Question

Is mean daily the most helpful response? What about variance from solar noon? What about extremes related to short events?

Plot available records:

```{r temp-plot, fig.width=9, fig.height=5}
ggplot(temp_daily, aes(x = date, y = mean_temp_c, color = site_name)) +
  geom_line() +
  labs(
    x = NULL,
    y = "Daily mean water temperature (°C)",
    color = "Site",
    title = "Available water-temperature records"
  ) +
  theme_minimal()
```

The Truckee River at Reno is useful for discharge and turbidity, but it does not currently provide the same continuous temperature record as Blackwood and Incline Creeks.

This is a real-world feature of environmental databases: **the best station depends on the question and variable.**

Optional nearby Truckee River temperature site:

```{r clark-temp, eval=FALSE}
clark_temp <- read_waterdata_continuous(
  monitoring_location_id = "USGS-10350500",
  parameter_code = "00010",
  time = c(start_date, end_date)
)
```

# Summarize turbidity carefully

Turbidity parameter `63680` is reported in FNU.

Rather than simply averaging every 15-minute value, first calculate daily summaries.

```{r daily-turbidity}
turbidity_daily <- sensor_clean %>%
  filter(parameter_code == "63680") %>%
  mutate(date = as.Date(datetime)) %>%
  group_by(site_name, date) %>%
  summarise(
    mean_turbidity_fnu = mean(value, na.rm = TRUE),
    median_turbidity_fnu = median(value, na.rm = TRUE),
    max_turbidity_fnu = max(value, na.rm = TRUE),
    n_obs = sum(!is.na(value)),
    .groups = "drop"
  )

turbidity_daily
```

Check completeness:

```{r turbidity-completeness}
turbidity_completeness <- turbidity_daily %>%
  group_by(site_name) %>%
  summarise(
    days_with_turbidity = n_distinct(date),
    first_day = min(date),
    last_day = max(date),
    .groups = "drop"
  )

turbidity_completeness
```

# Compare mean turbidity among watersheds

```{r turbidity-summary}
turbidity_summary <- turbidity_daily %>%
  group_by(site_name) %>%
  summarise(
    days_with_data = n(),
    mean_daily_turbidity_fnu = mean(mean_turbidity_fnu, na.rm = TRUE),
    median_daily_turbidity_fnu = median(mean_turbidity_fnu, na.rm = TRUE),
    maximum_daily_turbidity_fnu = max(max_turbidity_fnu, na.rm = TRUE),
    .groups = "drop"
  )

turbidity_summary
```

```{r turbidity-barplot, fig.width=9, fig.height=5}
ggplot(
  turbidity_summary,
  aes(
    x = reorder(site_name, mean_daily_turbidity_fnu),
    y = mean_daily_turbidity_fnu
  )
) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Mean daily turbidity (FNU)",
    title = "Mean turbidity during Water Year 2024"
  ) +
  theme_minimal()
```

# Turbidity through time

```{r turbidity-time, fig.width=10, fig.height=6}
ggplot(turbidity_daily, aes(x = date, y = mean_turbidity_fnu)) +
  geom_line() +
  facet_wrap(~ site_name, scales = "free_y", ncol = 1) +
  labs(
    x = NULL,
    y = "Daily mean turbidity (FNU)",
    title = "Turbidity is episodic"
  ) +
  theme_minimal()
```

### Questions

1.  Are the highest-turbidity days associated with high-flow periods?
2.  Would the annual mean or event maximum better describe sediment transport?
3.  Why might the answer depend on the ecological or management question?

# Put discharge and turbidity together

```{r join-q-turbidity}
q_turbidity <- q_daily_clean %>%
  select(site_name, date, discharge_cfs) %>%
  inner_join(
    turbidity_daily %>%
      select(site_name, date, mean_turbidity_fnu),
    by = c("site_name", "date")
  )

head(q_turbidity)
```

Plot turbidity against discharge:

```{r q-turbidity-plot, fig.width=9, fig.height=6}
ggplot(
  q_turbidity,
  aes(x = discharge_cfs, y = mean_turbidity_fnu)
) +
  geom_point(alpha = 0.5) +
  facet_wrap(~ site_name, scales = "free") +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    x = expression("Daily mean discharge (ft"^3*"/s; log scale)"),
    y = "Daily mean turbidity (FNU; log scale)",
    title = "Does turbidity increase with streamflow?"
  ) +
  theme_minimal()
```

> **Note:** Log scales cannot display zero values. Inspect zeros before deciding how to handle them.

# Summary table

```{r final-table}
final_summary <- annual_yield %>%
  left_join(
    turbidity_summary,
    by = "site_name"
  ) %>%
  arrange(desc(drainage_area_mi2))

final_summary
```

# Inspect the API request itself

The `dataRetrieval` functions construct API requests for us.

A daily mean discharge request follows this logic:

``` text
https://api.waterdata.usgs.gov/ogcapi/v0/collections/daily/items
    ?monitoring_location_id=USGS-10336660
    &parameter_code=00060
    &statistic_id=00003
    &time=2023-10-01/2024-09-30
```

The important parts are:

- **where?** monitoring location
- **what?** parameter code
- **which statistic?** mean
- **when?** date range

# Reproducibility

```{r session-info}
sessionInfo()
```
